### 2020-08-30 doc/manuscript/Dow_manuscript.Rmd
more work on introduction

As global atmospheric greenhouse gas levels are rising, and little meaningful policy is being implemented to combat this, the planet is expected to surpass the 2.5 degree C warming mark that the Paris accord pledged to avoid. This will have potentially negative implications for nearly every ecosystem on Earth, including forests. -importantce of forests Deciduous forests in the Northern Hemisphere have the potential to remove up to ## tons of carbon in the next ## years, and currently represent %% of the worlds carbon reserves. Trees remove CO2 from the atmosphere, and fix it into structural components like wood, roots, or leaves, through the process of photosynthesis. Current theory suggests that tree growth could benefit from global warming, as warmer early Spring time allows leaves to emerge earlier and increased CO2 concentrations increase plant water use efficiency. This increase in growth can slow the rate of global warming as CO2 sequestration is sped up, making forests key natural combatants in the fight against greenhouse gasses.

In recent decades, Spring has been arriving earlier as Spring temperatures are increasing. This leads to earlier leaf out in deciduous trees, meaning a potentially longer growing season (). Earlier Spring emergence has been shown to cause both positive and negative lagged effects on subsequent Autumn or Spring growth in European broad leaf deciduous and mixed forests, respectively (Crabbe et al, 2016). Even more, the timing of Spring warming seems to have a significant effect on leaf-out response, where warming events later in Spring were shown to impact leaf emergence timing more than earlier Spring warming (Freidl et al, 2014). These shifts in phenological timing are likely to affect woody productivity, carbon and nutrient cycling, and water use patterns in temperate forests. However, the fine-scale responses of temperate deciduous species to a warming Spring are not fully understood. Many predictions of phenological shifts like these are based on GIS, eddy-covarience measurements, or other broad scale observations of phenological timing, and do little to discern differences between individual species. In reality, we know little about the growth phenology of temperate deciduous species (see D’Orangeville et al., in revision for more info/ citations)—let alone how these are affected by an early spring.

To date, very little research has been done on the phenology of tree growth. While observations of leaf phenology have shown a decent relationship with NPP, understanding intraannual wood-growth phenology allows a finer-scale look at differences in climate sensitivity and growth responses between individual species. Recent results from D’Orangeville et al. (in prep) have shown that wood-type plays a significant role in the timing of leaf and wood phnology, as ring-porous begin growth earlier in the season compared to diffuse-porous, but both reached 75% total annual growth around the same time. This implies evolutionary differences in the mechanisms controlling growth of these species, which have the potential to expose each group to different climate stressors.

Temperate trees are projected to increase NPP and sequester greater amounts of CO2 as climate warms, to a point, but the contribution of individual species, and the implications of potentially chanign species compositions is less well known. This simplification is necessary at larger scales, as generalizations are needed to ensure models are not overly complex. However, the results here indicate that all tree species may not be responding equally to an earlier Spring. We suggest that efforts be made to examine regions on a smaller scale, investigating indiviudual species or wood-types’ reaction to historic warming to forecast future NPP. For example, at SCBI, the loss of ring-porous Fraxinus spp., the high prolification of the diffuse-porous Lirodendron Tulipera, and Decline of mature individuals along with browsing pressure on Quercus saplings could lead to a population boom of diffuse-porous species, making our forest less likely to increase growth in response to a warming early-Spring and thus, lessening the impact our forest has on the global carbon balance.

Dendrometer bands allow a look at indivudal trees’ intraannual growth. Using many measurements taken within a year allows the creation of an idealized growth curve using a 5 paramter logistic model (Sean’s model). From this, growth milestones and variables can be extracted from each year and compared to other years, effectivley simulating a controlled experiment where climate variables were manipulated.

To our knowledge, no research has been done exploring the relationship between intraannual growth patterns and early Spring temperatures. Here, we use 9 years of biweekly dendrometer band measurements for 4 species to characterize intraannual growth phenology within the FOrestGEO plot at SCBI, calculating for each tree in each year where data was available: the day of year (DOY) where 25, 50, and 75% annual growth was achieved; the maximum growth rate; the DOY where maximum growth rate was achieved; and the total growth. These variables were then compared to pre-season temperatures over the 9 year period of 2011-2019 to search for correlation between early-spring temps and growth phenology.

@@

Here, we … - use 10 years of biweekly dendro band measurements for # species to characterize seasonal growth phenology
•	combine these records with 110 years of tree-ring data for 12 species to examine how spring temperatures impact annual tree growth
•	examine the effects of an extremely early spring (2020) seasonal growth phenology and annual growth

@@

Study site and data
SCBI ForestGEO plot
-SCBI met-tower
Climate data was obtained from a meteorological tower in a clearing adjacent to, and within elevation range of the SCBI ForestGEO plot. This tower is part of the ForestGEO meteorological monitoring program. Temperature readings are taken every 5 minutes using a CR1000 datalogger.
9 years biweekly dendrometer measurements
Dendrometer band measurements were taken every two weeks for 9 years (After cleaning: 2011- 105 trees; 2012 - 102 trees; 2013 - 102 trees; 2014 - 149 trees; 2015 - 149 trees; 2016 - 149 trees; 2017 - 148 trees; 2018 - 146 trees; 2019 - 145 trees ). Data manually cleaned by visual inspection. Three classes of mistakes: 1. Error in measurement - weekly measurement was drastically different from previous week and following week. Mistakes were removed from the raw data before modeling. (## cases) 2. Band slip or stuck - measurements freeze and remain unchanged until sudden jump followed by normal growth pattern. Band slips were sometimes followed by measurements indicating the band became stuck; in these cases, the tree was thrown out for the corresponding year. In cases where the slip was followed by normal growth, the initial slip-point was removed, and the following points were shifted down to the pre-slip level. (## cases) 3. Other - cases where data was clearly wrong but with unknown causes. This seemed to happen mostly in 2011, when the program was just beginning. In cases where several measurements were an issue, they were simply removed. If there were no clear solutions to ‘fix’ the data, the entire year was removed from the analysis.

>>>>

As global atmospheric greenhouse gas levels are rising, and little meaningful policy is being implemented to combat this, our planet is expected to surpass the 2.5 degree C warming mark that the Paris accord pledged to avoid. The absence of human intervention increases our reliance on natural systems to soften the blow of climate change.

Forests are an extremely valuable ecosystem for many reasons. They house an astonishing 80% of terrestrial species and are capable of removing the greenhouse gas CO2 from the atmosphere and storing it in the trees or soil. Forests sequester about one-third of all anthropogenic CO2 and store ~45% of all terrestrial carbon (Bonan, 2008). However, their future under the current era of rapid change is uncertain (Friedlingstein et al, 2006). Reducing this uncertainty requires better understanding of the drivers of tree growth (Bugmann et al, 2019). Tree growth is sensitive to climatic factors like precipitation, temperature, and cloudiness in early growing season (Helcoski et al, 2019; Ciais, 2003; Goulden, 1996), as well as other physical factors including nitrogen content of soil, increasing atmospheric CO2, deposition of atmospheric pollutants in acid rain (Mathias & Thomas, 2018), and invasion by pests or pathogens. Understanding how growth is influenced by multiple factors at once is essential for developing accurate models to inform forest managers and policy makers about the role of future forests in climate change mitigation. CO2 concentrations in the atmosphere are increasing, and in a perfect scenario, trees and other plants would simply speed up growth in response to this, increasing photosynthetic efficiency in a process called CO2 fertilization (Strain and Cure, 1986; Huang et al., 2007). However, factors other than CO2 also limit tree growth, and seasonal temperature averages and precipitation patterns are changing concurrently. It is anticipated that, throughout much of the United States, future heat and drought may limit tree growth more strongly than it can be enhanced by elevated CO2 (Charney et al, 2016).

In recent decades, an increase in average early Spring temperatures has been attributed to climate change. As temperatures are increasing, some regions have seen earlier leaf out in deciduous trees, meaning a potentially longer growing season (Ibáñez et al., 2010; Jeong et al., 2011). The effects of this extension of the growing season on leaf phenology and NPP have been well studied (Way, 2011; ), but little work has been done to understand the impacts to intraannual stem-growth variables. Shifts in leaf phenology timing could be coupled with a change in stem-growth phenology, which are likely to affect woody productivity, carbon and nutrient cycling, and water use patterns in temperate forests. Thus, it is important to understand how the warming of Spring changes stem-growth phenology.

The timing of Spring warming has a significant effect on leaf phenology response. Warming events later in Spring were shown to impact leaf emergence timing more than earlier Spring warming (Freidl et al, 2014) and earlier leaf emergence has been shown to cause complex lagged effects on subsequent seasonal growth in European broad leaf deciduous and mixed forests (Crabbe et al, 2016). GIS source Ian sent. However, the fine-scale responses of temperate deciduous species to a warming Spring are not fully understood. Many predictions of phenological shifts like these are based on GIS, eddy-covarience measurements, or other broad scale observations of phenological timing, and do little to discern differences between individual species, where leaf and stem-growth phenology can be decoupled (D'Orangeville et al, in review). In reality, we know little about the growth phenology of temperate deciduous species (see D’Orangeville et al., in revision for more info/ citations)—let alone how these are affected by an early spring.

To date, very little research has been done on the phenology of stem-growth. Timing of intraannual growth is extremely improtant as it can dictate periods of highest drought sensitivity of NPP in temperate forests (Foster, 2014; ). For example, if a drought were to occur when trees were at their peak growth rates, it could decrease that year's NPP substantially, and subsequently, reduce the total amount of carbon sequestered creating a negative climate feedback loop. Recent results from D’Orangeville et al. (in review) have shown that wood-type plays a significant role in the timing of leaf and wood phenology, as ring-porous trees begin growth earlier in the season compared to diffuse-porous trees, but both reached 75% total annual growth around the same time. This implies evolutionary differences in the mechanisms controlling growth of these species, which have the potential to expose each group to different climate stresses.

It is difficult to map intraannual growth, or compare interspecific responses to Spring warming using leaf emergence timing alone. A solution to this is the use of dendrometer bands. Dendrometer bands allow a look at individual trees’ intraannual growth at a fine temporal scale. Using many measurements taken within a year, we can model the growth of individual trees and extract growth milestones and variables for interspecific, intraspecific, and temporal comparison (McMahon & Parker, 2014). This allows the determination of variable responses of all tree species within a single forest, meaning a better understanding of future growth under projected climate conditions. Here, we use 9 years of biweekly dendrometer band measurements for 4 species to characterize intraannual stem-growth phenology within the ForestGEO plot at SCBI, calculating for each tree in each year where data was available: the day of year (DOY) where 25, 50, and 75% annual growth was achieved; the maximum growth rate; the DOY where maximum growth rate was achieved; and the total growth. Correlation between these variables and pre-season temperatures over the 9 year study period (2011-2019) were assessed using a heirarchal bayesian model.

@@

@@

Study site and data
SCBI ForestGEO plot

ForestGEO is a global network of scientists and 71 forest plots dedicated to the long-term study of the world's forests. The ForestGEO plot at SCBI is a 26-ha forest located in Front Royal, Virginia. Every stem ≥1 cm is tagged, mapped, identified to species, and measured in diameter at breast height (DBH) every 5 years following standardized ForestGEO procedure (Condit, 1995). Within the plot, >500 trees have dendrometer bands which are measured at the start and end of each growing season, while a subset of ~150 are additionally measured every two weeks during the growing season.

Dendrometer band measurements were taken every two weeks for 9 years (After cleaning: 2011- 105 trees; 2012 - 102 trees; 2013 - 102 trees; 2014 - 149 trees; 2015 - 149 trees; 2016 - 149 trees; 2017 - 148 trees; 2018 - 146 trees; 2019 - 145 trees ). The data were manually cleaned by visual inspection. Three classes of mistakes: 1. Error in measurement - weekly measurement was drastically different from previous week and following week. Mistakes were removed from the raw data before modeling. (## cases) 2. Band slip or stuck - measurements freeze and remain unchanged until sudden jump followed by normal growth pattern. Band slips were sometimes followed by measurements indicating the band became stuck; in these cases, the tree was thrown out for the corresponding year. In cases where the slip was followed by normal growth, the initial slip-point was removed, and the following points were shifted down to the pre-slip level. (## cases) 3. Other - cases where data was clearly wrong but with unknown causes. This seemed to happen mostly in 2011, when the program was just beginning. In cases where several measurements were an issue, they were simply removed. If there were no clear solutions to ‘fix’ the data, the entire year was removed from the analysis.

Climate data used was obtained from the National Center for Environmental Information (NCEI) weather station located in Front Royal, Virginia. Average daily maximum temperature for 2011-2019 are from the Daily Summary dataset available on their website (https://www.ncdc.noaa.gov/cdo-web/datasets/GHCND/stations/GHCND:USC00443229/detail).
